Defining Greatness

### 2024-12-01 paper/paper.qmd
think final update for paper

---
title: "Defining Greatness"
subtitle: "Mahomes’ Path to Immortality vs. Brady’s Unmatched Legacy"
author: 
  - Alexander Guarasci
thanks: "Code and data are available at: https://github.com/AlexanderG123/nfl"
date: today
date-format: long
abstract: "This paper compares National Football League (NFL) quarterbacks Tom Brady and Patrick Mahomes, focusing on their statistical performances. Analyzing their respective career data, this paper uses predictive modelling to forecast Patrick Mahomes’ future. It finds that while both have impressive career statistics, ultimately, playoff victories, particularly Super Bowl wins, serve as a more definitive measure of greatness. By examining the performance of these players, the paper highlights how the NFL’s dynamic context impacts player evaluation and legacy assessment."
format: 
  pdf: 
    fig-pos: "H"
number-sections: false
bibliography: references.bib
---
```{r}
#| include: false
#| warning: false
#| message: false

@@

The paper is structured as follows: the first section provides an explanation of the data used in this paper, what it shows and where it comes from. The second section discusses the model used to predict the future of Patrick Mahomes’ career. The third section shows the results of the model. The final section discusses the implication of the model as well as provides more insight into the two players’ legacies.

@@

The analysis was conducted in R [@citeR], leveraging tools from the Tidyverse [@citetidyverse], Caret [@citecaret], Arrow [@citearrow], Knitr [@citeknitr], and Testthat [@citett] packages. The primary dataset included QB-specific data from every NFL game played since 2000, encompassing over 55 columns detailing various player performance metrics. Additional data sources were used for ESPN’s Quarterback Rating (QBR), which encapsulates QB efficiency across multiple dimensions of gameplay.

@@

# Total Passing Yards Plot
player_filtered %>%
  group_by(player_display_name) %>%
  summarise(
    total_passing_yards = sum(passing_yards, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = player_display_name, y = total_passing_yards, fill = player_display_name)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("red","blue")) +
  theme_minimal() +
  labs(
    title = "Total Passing Yards: Mahomes vs Brady",
    x = "Player",
    y = "Passing Yards",
    fill = ''
  )

```
[@fig-total-passing-yards] provides a comprehensive overview of the total passing yards achieved by Tom Brady and Patrick Mahomes. Tom Brady, with 102,614 total passing yards (including playoffs), significantly outpaces Patrick Mahomes, who has 35,963 yards as of November 21, 2024. This disparity is largely due to Brady’s 381 career games compared to Mahomes’ 124 games. What stands out is Brady’s ability to maintain high passing numbers well into the later stages of his career, demonstrating longevity and consistency. On the other hand, Mahomes’ current total is remarkable given the shorter duration of his career, highlighting his rapid ascent and dominance in the modern NFL.

@@

# Total Rushing Yards Plot
player_filtered %>%
  group_by(player_display_name) %>%
  summarise(
    total_rushing_yards = sum(rushing_yards, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = player_display_name, y = total_rushing_yards, fill = player_display_name)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("red","blue")) +
  theme_minimal() +
  labs(
    title = "Total Rushing Yards: Mahomes vs Brady",
    x = "Player",
    y = "Rushing Yards",
    fill = ''
  )
```
[@fig-total-rushing-yards] shows an interesting contrast in rushing yards between the two quarterbacks. Despite playing over three times as many games as Mahomes, Tom Brady falls significantly short, with Mahomes amassing more than twice as many rushing yards. This reflects a broader evolution in the NFL, where modern quarterbacks are often required to be dual threats, capable of both passing and running effectively. Players like Mahomes exemplify this shift, adapting to the athletic demands of today’s game, unlike Brady, who represents the traditional pocket-passing archetype of previous eras.

@@

# Passing Touchdowns per Season Plot
player_data %>%
  group_by(player_display_name, season) %>%
  summarise(
    total_passing_tds = sum(passing_tds, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = season, y = total_passing_tds, color = player_display_name)) +
  geom_line() +
  theme_minimal() +
  labs(
    title = "Passing Touchdowns per Season",
    x = "Season",
    y = "Total Passing Touchdowns",
    colour = "Player"
  )

```
[@fig-passing-tds-per-season] illustrates the passing touchdowns per season for both players. Brady’s peak in 2007 with the New England Patriots—arguably the most dominant offense in NFL history—stands out prominently. Mahomes’ first season as a starter in 2018 is also noteworthy, with 53 touchdowns and just 12 interceptions. This extraordinary season set the stage for Mahomes’ rise as one of the game’s best quarterbacks, though it ultimately ended in a playoff loss to Brady. The graph underscores how both players have reached incredible heights in their respective careers.
```{r}
#| label: fig-passing-yard-distribution
#| fig-cap: Passing Yard Distribution for Mahomes and Brady
#| echo: false
#| warning: false
#| message: false

player_data %>%
  ggplot(aes(x = player_display_name, y = passing_yards, fill = player_display_name)) +
  geom_violin(alpha = 0.7,trim = FALSE) +
  theme_minimal() +
  scale_fill_manual(values = c("Mahomes" = "blue", "Brady" = "red")) +
  labs(
    title = "Passing Yards Distribution",
    x = "Player",
    y = "Passing Yards",
    fill = "Player"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    legend.position = "none"
  )
```
[@fig-passing-yard-distribution] examines the distribution of passing yards per game for each player. While both quarterbacks exhibit similar averages, Brady’s larger sample size, due to his longer career, shows higher highs and lower lows. This graph captures the consistency of both players but also highlights the significance of longevity with respect to great and poor performances. Mahomes’ distribution reflects his efficiency and performance within a shorter time frame.

@@

# QBR Comparison Plot
ggplot(qbr_filtered, aes(x = season_type, y = qbr_total, fill = name_display)) +
  geom_boxplot() +
  theme_minimal() +
  labs(
    title = "QBR Comparison by Season Type (Regular vs. Playoffs)",
    x = "",
    y = "QBR",
    fill = 'Player'
  )
```

@@

# Model

@@

- **completions**: The number of successful passes thrown.
- **attempts**: The total number of pass attempts.
- **passing_yards**: The total yards gained from passing attempts.
- **passing_tds**: The total number of passing touchdowns.
- **interceptions**: The total number of interceptions thrown.
- **carries**: The number of rushing attempts.
- **rushing_yards**: The total rushing yards gained.
- **rushing_tds**: The total number of rushing touchdowns.

The model used in this analysis is a multiple linear regression model. We choose linear regression because it is a simple yet effective approach for modeling continuous outcomes and provides easy interpretability.

@@

$$
Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots + \beta_n X_n + \epsilon
$$

@@

We also calculated the Root Mean Squared Error (RMSE) for each predicted variable in the test set as an evaluation metric. RMSE is a commonly used metric to measure the accuracy of regression models, and it is calculated as:

$$
RMSE = \sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y}_i)^2}
$$

@@

## Limitiations

While the RMSE values suggest some error in the predictions, this is expected due to the limited amount of available data. The performance of the model could likely improve with a larger and more balanced dataset that includes a broader range of game conditions. Additionally, while we used Tom Brady’s statistics as a proxy for Mahomes, differences in their playing styles and team dynamics could also contribute to the prediction error.

The linear regression model assumes that the relationships between the predictors and the outcomes are linear, which may not always be the case in real-world scenarios. Further exploration of alternative models, such as non-linear regression or machine learning techniques like decision trees, could be considered for improving predictions.

@@

# Results
 [@tbl-brady-mahomes-career-stats] compares Tom Brady’s career statistics with Patrick Mahomes’ projected lifetime statistics. Tom Brady, with 8,954 completions and 13,973 attempts, has significantly more career completions and attempts than Mahomes, who is projected to finish with 7,763 completions and 11,925 attempts. Brady’s total passing yards of 102,616 far exceed Mahomes’ projected 89,160, highlighting Brady’s longevity and consistent performance. Similarly, Brady has thrown 738 passing touchdowns, whereas Mahomes is projected to finish with 654. Brady also has more interceptions (252 compared to Mahomes’ 216). These differences underscore Brady’s long career and high volume of passing, which has contributed to his unrivaled career achievements.

@@

# Generate the table using tinytable
tt(final_table, width = 1.5) |> style_tt( fontsize = 1)
```

@@

# Discussion

>>>>

---
title: "Defining Greatness"
subtitle: "Mahomes’ Path to Immortality vs. Brady’s Unmatched Legacy"
author: 
  - Alexander Guarasci
thanks: "Code and data are available at: https://github.com/AlexanderG123/nfl"
date: today
date-format: long
abstract: "This paper compares National Football League (NFL) quarterbacks Tom Brady and Patrick Mahomes, focusing on their statistical performances. Analyzing their respective career data, this paper uses predictive modelling to forecast Patrick Mahomes’ future. It finds that while both have impressive career statistics, ultimately, playoff victories, particularly Super Bowl wins, serve as a more definitive measure of greatness. By examining the performance of these players, the paper highlights how the NFL’s dynamic context impacts player evaluation and legacy assessment."
format: 
  pdf: 
    fig-pos: "H"
number-sections: true
bibliography: references.bib
---
```{r}
#| include: false
#| warning: false
#| message: false

@@

The paper is structured as follows: @sec-data provides an explanation of the data used in this paper, what it shows and where it comes from. @sec-model discusses the model used to predict the future of Patrick Mahomes’ career. @sec-results shows the results of the model. @sec-discussion discusses the implication of the model as well as provides more insight into the two players’ legacies.

@@

The analysis was conducted in R [@citeR], leveraging tools from the Tidyverse [@citetidyverse], Caret [@citecaret], Arrow [@citearrow], Knitr [@citeknitr], and Testthat [@citett] packages. The code was styled using [@citestyler] The primary dataset included QB-specific data from every NFL game played since 2000, encompassing over 55 columns detailing various player performance metrics. Additional data sources were used for ESPN’s Quarterback Rating (QBR), which encapsulates QB efficiency across multiple dimensions of gameplay.

@@

# Total Passing Yards Plot
player_filtered %>%
  group_by(player_display_name) %>%
  summarise(
    total_passing_yards = sum(passing_yards, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = player_display_name, y = total_passing_yards, fill = player_display_name)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("red", "blue")) +
  theme_minimal() +
  labs(
    title = "Total Passing Yards: Mahomes vs Brady",
    x = "Player",
    y = "Passing Yards",
    fill = ""
  )
```
[@fig-total-passing-yards] provides a comprehensive overview of the total passing yards achieved by Tom Brady and Patrick Mahomes. Tom Brady, with 102,614 total passing yards (including playoffs), significantly outpaces Patrick Mahomes, who has 35,963 yards as of November 21, 2024. This disparity is largely due to Brady’s 381 career games compared to Mahomes’ 124 games. What stands out is Brady’s ability to maintain high passing numbers well into the later stages of his career, demonstrating longevity and consistency. On the other hand, Mahomes’ current total is remarkable given the shorter duration of his career, highlighting his rapid ascent and dominance in the modern NFL.

@@

# Total Rushing Yards Plot
player_filtered %>%
  group_by(player_display_name) %>%
  summarise(
    total_rushing_yards = sum(rushing_yards, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = player_display_name, y = total_rushing_yards, fill = player_display_name)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("red", "blue")) +
  theme_minimal() +
  labs(
    title = "Total Rushing Yards: Mahomes vs Brady",
    x = "Player",
    y = "Rushing Yards",
    fill = ""
  )
```
[@fig-total-rushing-yards] shows an interesting contrast in rushing yards between the two quarterbacks. Despite playing over three times as many games as Mahomes, Tom Brady falls significantly short, with Mahomes amassing more than twice as many rushing yards. This reflects a broader evolution in the NFL, where modern quarterbacks are often required to be dual threats, capable of both passing and running effectively. Players like Mahomes exemplify this shift, adapting to the athletic demands of today’s game, unlike Brady, who represents the traditional pocket-passing archetype of previous eras.

@@

# Passing Touchdowns per Season Plot
player_data %>%
  group_by(player_display_name, season) %>%
  summarise(
    total_passing_tds = sum(passing_tds, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = season, y = total_passing_tds, color = player_display_name)) +
  geom_line() +
  theme_minimal() +
  labs(
    title = "Passing Touchdowns per Season",
    x = "Season",
    y = "Total Passing Touchdowns",
    colour = "Player"
  )
```
[@fig-passing-tds-per-season] illustrates the passing touchdowns per season for both players. Brady’s peak in 2007 with the New England Patriots—arguably the most dominant offense in NFL history—stands out prominently. Mahomes’ first season as a starter in 2018 is also noteworthy, with 53 touchdowns and just 12 interceptions. This extraordinary season set the stage for Mahomes’ rise as one of the game’s best quarterbacks, though it ultimately ended in a playoff loss to Brady. The graph underscores how both players have reached incredible heights in their respective careers.
```{r}
#| label: fig-passing-yard-distribution
#| fig-cap: Passing Yard Distribution for Mahomes and Brady
#| echo: false
#| warning: false
#| message: false

player_data %>%
  ggplot(aes(x = player_display_name, y = passing_yards, fill = player_display_name)) +
  geom_violin(alpha = 0.7, trim = FALSE) +
  theme_minimal() +
  scale_fill_manual(values = c("Mahomes" = "blue", "Brady" = "red")) +
  labs(
    title = "Passing Yards Distribution",
    x = "Player",
    y = "Passing Yards",
    fill = "Player"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    legend.position = "none"
  )
```
[@fig-passing-yard-distribution] examines the distribution of passing yards per game for each player. While both quarterbacks exhibit similar averages, Brady’s larger sample size, due to his longer career, shows higher highs and lower lows. This graph captures the consistency of both players but also highlights the significance of longevity with respect to great and poor performances. Mahomes’ distribution reflects his efficiency and performance within a shorter time frame.

@@

# QBR Comparison Plot
ggplot(qbr_filtered, aes(x = season_type, y = qbr_total, fill = name_display)) +
  geom_boxplot() +
  theme_minimal() +
  labs(
    title = "QBR Comparison by Season Type (Regular vs. Playoffs)",
    x = "",
    y = "QBR",
    fill = "Player"
  )
```

@@

# Model {#sec-model}

@@

- **completions**: The number of successful passes thrown.
- **attempts**: The total number of pass attempts.
- **passing yards**: The total yards gained from passing attempts.
- **passing tds**: The total number of passing touchdowns.
- **interceptions**: The total number of interceptions thrown.
- **carries**: The number of rushing attempts.
- **rushing yards**: The total rushing yards gained.
- **rushing tds**: The total number of rushing touchdowns.

The model used in this analysis is a multiple linear regression model (@eq-lm). We choose linear regression because it is a simple yet effective approach for modeling continuous outcomes and provides easy interpretability.

@@

$$
Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots + \beta_n X_n + \epsilon
$$ {#eq-lm}

@@

We also calculated the Root Mean Squared Error (RMSE) for each predicted variable in the test set as an evaluation metric. RMSE (@eq-rmse) is a commonly used metric to measure the accuracy of regression models, and it is calculated as:

$$
RMSE = \sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y}_i)^2}
$$ {#eq-rmse}

@@

## Assumptions and Limitations

The model makes several key assumptions that are important for interpreting the predictions:

1. **Similarity of Player Performance**: The model assumes that Patrick Mahomes' future performance will be similar to Tom Brady’s past performance, using Brady's career as a proxy. This assumption may not hold as Mahomes' playing style and team dynamics significantly differ from Brady's.

2. **Linear Relationships**: The model assumes that the relationships between the predictors (game statistics such as completions, passing yards, etc.) and the outcomes are linear. This may not always be the case in practice, as player performance could be influenced by more complex, non-linear interactions.

3. **Constant Game Conditions**: The model assumes that game conditions, including factors such as opposing defenses and weather conditions, remain relatively consistent over time. However, these factors could vary and influence player performance in ways not captured by the model.

4. **Prediction for 200 Games**: The model assumes that Mahomes' performance will follow a similar trajectory over the next 200 games as it has in the past. This assumption may not account for potential career milestones, injuries, or other events that could disrupt this pattern.

5. **Uniform Dataset**: The model assumes that the dataset of Tom Brady’s career is representative of typical game conditions and does not contain significant biases or outliers that would skew the predictions.

Ultimately, the model may be appropriate in scenarios where general trends in quarterback performance are needed for broad projections, such as for media analysis or high-level discussions of player comparisons. It can offer insights into how Mahomes might perform in a typical game based on historical data from a similar elite quarterback. This can be useful for estimating the overall trajectory of Mahomes’ career and projecting his future career stats across a large sample of games.

However, the model may not be appropriate for precise, game-by-game predictions, especially when specific game conditions or player-specific factors (such as injury status or team dynamics) are likely to vary. For example, using this model to predict Mahomes’ performance in a critical playoff game or during a season affected by significant roster changes could lead to inaccurate or misleading conclusions. Additionally, the linear nature of the model means it may fail to account for more complex interactions that influence a quarterback’s performance in real-world scenarios.

In summary, while the model provides a reasonable estimation of Mahomes’ future statistics based on Brady’s historical data, it should be used cautiously in contexts where precision is critical or where Mahomes’ unique playing style and career circumstances may diverge from Brady’s past performance.

@@

# Results {#sec-results}
 [@tbl-brady-mahomes-career-stats] compares Tom Brady’s career statistics with Patrick Mahomes’ projected lifetime statistics. Tom Brady, with 8,954 completions and 13,973 attempts, has significantly more career completions and attempts than Mahomes, who is projected to finish with 7,763 completions and 11,925 attempts. Brady’s total passing yards of 102,616 far exceed Mahomes’ projected 89,160, highlighting Brady’s longevity and consistent performance. Similarly, Brady has thrown 738 passing touchdowns, whereas Mahomes is projected to finish with 654. Brady also has more interceptions (252 compared to Mahomes’ 216). These differences underscore Brady’s long career and high volume of passing, which has contributed to his unrivaled career achievements.

@@

# Generate the table using tinytable
tt(final_table, width = 1.5) |> style_tt(fontsize = 1)
```

@@

# Discussion {#sec-discussion}
